Scalar Feedback based Joint Time-Frequency Precoder interpolation for MIMO-OFDM Systems

### 2019-12-25 paper/main.tex
Minor fix
Minor intro rewrite

The use of multiple antennas at the transmitter and receiver (MIMO)
can significantly enhance the performance of wireless systems.  A
typical step to get this performance enhancements is estimation of
channel state information (CSI).  If transmitter also has the
knowledge of CSI, it can allocate resources more efficiently, both to
improve data rates and to reduce BER~\cite{love2008overview}.  To this
end, it is necessary to efficiently feed back CSI from the receiver to
the transmitter. In wireless systems that use orthogonal frequency
division multiplexing (OFDM), the feedback needs to be provided to the
transmitter across all subcarriers and in each OFDM frame, which
places a large feedback demand. However, since the CSI varies
gradually in both time and frequency, effective techniques to reduce
feedback overheads exist, using frequency domain interpolation and
temporal prediction [CITE?]. In particular, knowledge of the precoding
matrices (that are typically the orthonormal right singular vectors of
CSI matrix) is essential to achieve the link capacity, as well as to
parallelize the channel. In this paper, we focus on the problem of
tracking the precoders of a MIMO-OFDM system across time and frequency
for efficient CSI feedback by tracking a parametrized version of the
precoding matrices.

The problem of effective quantization and feedback of precoders for
MIMO-OFDM systems has attracted significant attention in recent
years. Many approaches involve codebook design over the Grassmann
manifold~\cite{mondal2007quantization,schwarz2013adaptive,5946308},
the Stiefel manifold \cite{6891198,Gupt1905:Predictive} and the Flag
manifold~\cite{pitaval2013codebooks}, wherein the manifold structure
of the precoders is used to obtain tangents for predictive
quantization and interpolation across time and frequency.
An alternate approach is to parametrize the precoding matrices using
scalar parameters. The space of unitary matrices can be parameterized into
independent scalar parameters that correspond to Givens rotations and
Householder transformations of the unitary precoding matrices
\cite{4114278,4556174}.

While the manifold based methods have been shown to be very effective
for precoder feedback, they involve operations over higher dimensional
manifolds. In particular, optimization and interpolation is
complicated, especially for manifolds where geodesic paths for
interpolation may be difficult to obtain, especially (don't repeat
especially again) for joint time-frequency predictive coding.  In
addition to being analytically more tractable, the scalar
parametrization methods have also been implemented in 802.11ac
standards \cite{porat2013improved}. Past work in utilizing scalar
parametrization for precoder feedback have utilized adaptive delta
modulation effectively track precoding matrices both in the
time~\cite{4114278} and frequency domains~\cite{4556174}, albeit
separately. In this paper, we propose a method for exploiting the
joint time-frequency correlation of the precoder's scalar parameters
on the transmitter using CSI feedback.  Typical approaches to
predictive quantization and interpolation do not exploit the joint
time-frequency correlation to enhance the precoder estimate at the
transmitter. However, exploiting the correlations jointly not only
reduces the feedback requirement, but also enhances the precoder
reconstruction accuracy significantly.

@@

\section{System Model}
For a MIMO-OFDM system with $N_T$ transmitter and $N_R$ receiver
antennas, the channel for the $i$-th subcarrier at time instant $t$
can be modelled as a flat-fading MIMO channel. This flat-fading MIMO
channel can be written as
\label{section2}
\begin{equation}
\by_{i,t} = \bH_{i,t}\tilde{\bV}_{i,t} \bx_{i,t}+ {\bf\eta}_{i,t}
\end{equation}
where $\bH_{i,t} \in \mathbb{C}^{N_R \times N_T}$ is the channel
matrix, $\bx_{i,t} \in \mathbb{C}^{N_d}$ is the input data vector,
$\by_{i,t} \in \mathbb{C}^{N_R}$ is the output vector,
$\tilde{\bV}_{i,t} \in \mathbb{C}^{N_T \times N_R}$ is the unitary
precoder, and $\eta$ is the complex additive white Gaussian noise
distributed as ${\mathcal{CN}}(0,I_{N_R})$. We consider the case where
$N_R < N_T$, and the data vector size $N_d = N_R$.  The ``thin'' SVD
(that keeps only those right and left singular vectors that correspond
to non-zero singular values) of $\bH$ is given by
$\bH_{i,t} = \bU_{i,t} \Sigma_{i,t} \bV_{i,t}^{H}$, where
$\bU_{i,t} \in \mathbb{C}^{N_R \times N_R}$,
$\Sigma_{i,t} \in \mathbb{C}^{N_R \times N_R}$ and
$\bV_{i,t} \in \mathbb{C}^{N_T \times N_R}$. $\bU_{i,t}$ and
$\bV_{i,t}$ are matrices whose columns are orthonormal, and
$\Sigma_{i,t}$ contains the singular values
$\sigma_1 \geq \ldots \geq \sigma_{N_R} > 0$ of $\bH_{i,t}$. The
entries of $\bH_{i,t}$ are assumed to be distributed
i.i.d. $\mathcal{CN}(0,1)$. To obtain a precoder at the
transmitter, we need to quantize and feed back information about
$\bV_{i,t}$ for each subcarrier. However, the frequency coherence and
temporal correlations would reduce the overall feedback requirement,
as discussed in subsequent sections.

>>>>

Multiple-input multiple-output (MIMO) based wireless communication
systems that use orthogonal frequency division multiplexing (OFDM)
have been adapted in many modern wireless systems owing to their
flexibility and ability to use the wideband channels effectively. One
key requirement to utilize such channels effectively is for the
transmitter to possess channel state information (CSI), since that
permits channel parallelization for low-complexity decoders as well as
the use of capacity achieving transmission
strategies~\cite{love2008overview}. This primarily translates to the
generation of unitary precoding matrices at the transmitter. However,
feeding back CSI from the receiver to the transmitter requires large
feedback overhead, since a precoder needs to be fed back for each OFDM
subcarrier within each coherence interval. Typical past approaches
have used manifold based approaches that exploit temporal and
frequency correlation to reduce feedback overheads. Wireless standards
have largely employed parametric quantization that uses Givens
rotations based quantization~\cite{lou2013comparison,5671092}, though these do
not involve any refinement of the estimates using channel
correlations. In this paper, we propose a joint time-frequency
interpolation approach that builds upon the parametric quantization
techniques used by standards that manages to outperform typical
manifold based quantization techniques.

Most approaches for effective quantization and feedback of precoders
for MIMO-OFDM systems involve codebook design over the Grassmann
manifold~\cite{mondal2007quantization,schwarz2013adaptive,5946308},
the Stiefel manifold \cite{6891198,Gupt1905:Predictive} and the Flag
manifold~\cite{pitaval2013codebooks}, wherein the manifold structure
of the precoders is used to obtain tangents for predictive
quantization and interpolation across time and frequency.  An
alternate approach is to parametrize the precoding matrices using
scalar parameters. The space of unitary matrices can be parameterized
into independent scalar parameters that correspond to Givens rotations
and Householder transformations of the unitary precoding matrices
\cite{4114278,4556174}.

While the manifold based methods have been shown to be very effective
for precoder feedback, they involve operations over higher dimensional
manifolds. In particular, optimization and interpolation is
complicated, especially for manifolds where geodesic paths for
interpolation may be difficult to obtain, especially (don't repeat
especially again) for joint time-frequency predictive coding.  In
addition to being analytically more tractable, the scalar
parametrization methods have also been implemented in 802.11ac
standards \cite{lou2013comparison}. Past work in utilizing scalar
parametrization for precoder feedback have utilized adaptive delta
modulation effectively track precoding matrices both in the
time~\cite{4114278} and frequency domains~\cite{4556174}, albeit
separately. In this paper, we propose a method for exploiting the
joint time-frequency correlation of the precoder's scalar parameters
on the transmitter using CSI feedback.  Typical approaches that feed
back scalar precoder parameters problem do not exploit the joint
time-frequency correlation to enhance the precoder estimate at the
transmitter. We show that exploiting correlations jointly not only
reduces the feedback requirement, but also enhances the precoder
reconstruction accuracy significantly.

@@

\section{System Model}
For a MIMO-OFDM system with $N_T$ transmitter and $N_R$ receiver
antennas, the channel for the $i$-th subcarrier at time instant $t$
can be modelled as a flat-fading MIMO channel. This flat-fading MIMO
channel can be written as
\label{section2}
\begin{equation}
\by_{i,t} = \bH_{i,t}\tilde{\bV}_{i,t} \bx_{i,t}+ \bf{\eta_{i,t}}
\end{equation}
where $\bH_{i,t} \in \mathbb{C}^{N_R \times N_T}$ is the channel
matrix, $\bx_{i,t} \in \mathbb{C}^{N_d}$ is the input data vector,
$\by_{i,t} \in \mathbb{C}^{N_R}$ is the output vector,
$\tilde{\bV}_{i,t} \in \mathbb{C}^{N_T \times N_R}$ is the unitary
precoder, and $\eta$ is the complex additive white Gaussian noise
distributed as ${\mathcal{CN}}(0,I_{N_R})$. We consider the case where
$N_R < N_T$, and the data vector size $N_d = N_R$.  The ``thin'' SVD
(that keeps only those right and left singular vectors that correspond
to non-zero singular values) of $\bH$ is given by
$\bH_{i,t} = \bU_{i,t} \Sigma_{i,t} \bV_{i,t}^{H}$, where
$\bU_{i,t} \in \mathbb{C}^{N_R \times N_R}$,
$\Sigma_{i,t} \in \mathbb{C}^{N_R \times N_R}$ and
$\bV_{i,t} \in \mathbb{C}^{N_T \times N_R}$. $\bU_{i,t}$ and
$\bV_{i,t}$ are matrices whose columns are orthonormal, and
$\Sigma_{i,t}$ contains the singular values
$\sigma_1 \geq \ldots \geq \sigma_{N_R} > 0$ of $\bH_{i,t}$. The
entries of $\bH_{i,t}$ are assumed to be distributed
i.i.d. $\mathcal{CN}(0,1)$. To obtain a precoder at the
transmitter, we need to quantize and feed back information about
$\bV_{i,t}$ for each subcarrier. However, the frequency coherence and
temporal correlations would reduce the overall feedback requirement,
as discussed in subsequent sections.
